Exploring Rural Shrink Smart Through Guided Discovery Dashboards

### 2022-07-31 SDSS Journal Paper/Paper Draft.Rmd
7.31.2022

The Smart and Connected Community (SCC) dashaboard data are primarily assembled from  \url{data.iowa.gov}, with some additional datasets assembled from federal and private sources. Most of these data sets are collected at a town/city or county spatial resolution, requiring us to carefully join data to ensure that these differences are respected while collating relevant information at the city level. In addition to the more commonly available statistics derived from e.g. the census and American Community Survey, \url{data.iowa.gov} contains several unique data sets, including local liquor sales, school building locations, town budgets and expenditures, hospital beds, Medicaid reimbursements, and other details that may provide information about local quality of life.

@@

Data collected from the Index of Relative Rurality (IRR) [@Rural_classification]  were used in the SCC dashboard to help classify the towns. The Index of Relative Rurality (IRR) is a continuous, threshold-free, and unit-free measure of rurality. It is an alternative to the traditional discrete threshold-based classifications.The IRR ranges between 0 (low level of rurality, i.e., urban) and 1 (most rural). Four steps are involved in its design:
\begin{enumerate}
\item Identifying the dimensions of rurality: population size, density, remoteness, and built-up area.
\item Selecting measureable variables to adequately represent each dimension:
    \begin{itemize}
        \item Size: logarithm of population size
        \item Density: logarithm of population density.
        \item Remoteness: network distance.
        \item Built-up area: urban area (as defined by the US Census Bureau) as a percentage of total land area.
    \end{itemize}
\item Re-scaling the variables onto bounded scales that range from 0 to 1.
\item Selecting a link function: unweighted average of the four re-scaled variable.
\end{enumerate}

@@

\svp{XXX Denise, can you follow up with the initial design section and explain the 2nd design and approach? Why did we change things? How did those changes impact feedback from users? E.g. wanting to show the whole range of values in the PCP vs. being visually overwhelmed. Incorporate the user feedback into the narrative. XXX}

\db{In the initial design, we included a map for each vital service, this initially created a lag for the users' experience. As a result, we cached map directions from OSRM for each service in our database, which drastically} \svp{reduced} \db{the response time for the user. Our initial design \svp{did not naturally focus} the user's eye on the most important parts of the dashboard; the redesign allowed for a cleaner \svp{flow} from the top to the bottom. }

\db{In addition to the timing due to the map loading slowly, we added the vital statistics at the county level to allow for a more robust understanding of the town and it's surroundings. The rurality index \svp{provided a better classification} and the USDA sources allowed for the town to understand \svp{the impact of} the closest major city \svp{due to commuting for work and shopping at larger stores not available within the town}.}

\svp{We also modified the parallel coordinate plots in several ways:}

@@

\svp{While not all of these modifications were well received in our second round of user testing, the changes did incrementally move the dashboard display towards our goal of allowing users to explore the data and engage with it. We continued to be surprised with how well users reacted to the parallel coordinate plots, which we initially thought might be too abstract for users unfamiliar with multivariate data displays, but the ability to compare towns across multiple dimensions and examine the similarities and differences between their approaches to different services seemed to be intuitive for users once they understood that each vertical axis was a different variable.}

\section{ Discussion}
How did the philosophy work out? What worked and didn't (overall)? People didn't seem to like having their community facts to explore from - they already knew that information - so what should we replace it with? Collaborator suggestions included reporting capability and data stories (narratives from the communities), but neither one really inspires exploratory data analysis. What we need to do is provide enough interactivity to allow people to select interesting data sets and use that to compare. We also should provide much, much more user support for how to interact with and interpret the app and results.

\section{ Future Work }

Incorporating other dashboards which had slightly different focuses (99 vs 990 towns, general data display, etc.). Transition to a multi-tab dashboard which allows for more specific task-oriented exploration and evaluation, rather than broad EDA and discovery.
Discussion of the difficulty of working with researchers from other disciplines, not because of convention but because they don't know what is possible with EDA and are unfamiliar with all but goal-directed data analysis.

\section {Conclusions}
Specific takeaways - visualization research (broad) vs. dashboard design (very problem specific) is missing a critical set of research describing the specific and trying to make general suggestions as to how to proceed with the design process. 
Helpful hints - showing a collection of very different apps to train collaborators on what is possible works well to empower collaborators to help brainstorm.
Test early, test often...

@@

\subsection{Visualization Revisited}
In our statistical graphs, we used Parallel Coordinate Plots (PCPs) to allow the user to look at a variety of variables, while the Guided Discovery Learning was used to makes sure that the users were informed of the information that was presented, it did not allow for the town leaders any more ease in understanding. As a result, we decided to look for easier ways to display data, such as boxplots and barplots. This direction will be limited but may be useful foor the sake of ease on the town leaders.

@@

\subsection{Feedback}
We will continue to engage with twon leaders to ensure that we are developing a product that will help them make better data-driven decisions that are useful to increasing their quality of life, which may lend itself to help less shrinkage in these rural areas.

>>>>

The Smart and Connected Community (SCC) dashboard data are primarily assembled from  \url{data.iowa.gov}, with some additional datasets assembled from federal and private sources. Most of these data sets are collected at a town/city or county spatial resolution, requiring us to carefully join data to ensure that these differences are respected while collating relevant information at the city level. In addition to the more commonly available statistics derived from e.g. the census and American Community Survey, \url{data.iowa.gov} contains several unique data sets, including local liquor sales, school building locations, town budgets and expenditures, hospital beds, Medicaid reimbursements, and other details that may provide information about local quality of life.

@@

Data collected from the Index of Relative Rurality (IRR) [@Rural_classification]  were used in the SCC dashboard to help classify the towns. The Index of Relative Rurality (IRR) is a continuous, threshold-free, and unit-free measure of rurality. It is an alternative to the traditional discrete threshold-based classifications.The IRR ranges between 0 (low level of rurality, i.e., urban) and 1 (most rural). Four steps are involved in its design:
\begin{enumerate}
\item Identifying the dimensions of rurality: population size, density, remoteness, and built-up area.
\item Selecting measurable variables to adequately represent each dimension:
    \begin{itemize}
        \item Size: logarithm of population size
        \item Density: logarithm of population density.
        \item Remoteness: network distance.
        \item Built-up area: urban area (as defined by the US Census Bureau) as a percentage of total land area.
    \end{itemize}
\item Re-scaling the variables onto bounded scales that range from 0 to 1.
\item Selecting a link function: unweighted average of the four re-scaled variable.
\end{enumerate}

@@

In the initial design, we included a map for each vital service, this initially created a lag for the users' experience. As a result, we cached map directions from OSRM for each service in our database, which drastically reduced the response time for the user. Our initial design did not naturally focus the user's eye on the most important parts of the dashboard; the redesign allowed for a cleaner flow from the top to the bottom.

In addition to the timing due to the map loading slowly, we added the vital statistics at the county level to allow for a more robust understanding of the town and it's surroundings. The rurality index provided a better classification and the USDA sources allowed for the town to understand the impact of the closest major city due to commuting for work and shopping at larger stores not available within the town.

We also modified the parallel coordinate plots in several ways:

@@

While not all of these modifications were well received in our second round of user testing, the changes did incrementally move the dashboard display towards our goal of allowing users to explore the data and engage with it. We continued to be surprised with how well users reacted to the parallel coordinate plots, which we initially thought might be too abstract for users unfamiliar with multivariate data displays, but the ability to compare towns across multiple dimensions and examine the similarities and differences between their approaches to different services seemed to be intuitive for users once they understood that each vertical axis was a different variable.

\section{ Discussion}
How did the philosophy work out?  
\db{The design philosophy worked well in the ways to promote guided discovery learning for the user. The blocks that were created helped create a way for the user to scroll to the next section. This framework in the design allowed for the highlighted pieces of the town in generally. This design allowed the user to see the PCP after understanding how the dashboard allowed for their town to centered. The PCP in the lower part of the dashboard allowed for the user to see the plot and adjust to the fact that they could add more towns to the plot.}

What worked and didn't (overall)?
\db{The idea to use the guided discovery methodology created a lack of understanding in the plot for a novice user. The need to explain a visual means that the user is more likely to not use the plot effectively. Our hope was that the PCP would be a bit more "self-explanatory" than it was initially. The number of helpful hints clutters the UI and may be a distraction for a particular user. }

People didn't seem to like having their community facts to explore from - they already knew that information - so what should we replace it with? Collaborator suggestions included reporting capability and data stories (narratives from the communities), but neither one really inspires exploratory data analysis. What we need to do is provide enough interactivity to allow people to select interesting data sets and use that to compare

\db{Many small towns don't like to see that their information is being displayed publicly or used to be compared to another city. This could a huge barrier for our resarch and work. To make sure that this does not become a problem, we could make sure that when allowing the user to compare towns in the PCP, we make sure to de-idenitify the towns in the UI. For example, the town user selects similiar population size, the each town that is populated will have a collection of ten similiar towns using a label, such as "Town 1", "Town 2", etc. }
\db{Along with the de-idenitifying other towns within the UI, using a login system for each town will allow for the town to feel as if their data is being protected. This login system can then be used as a filter to setup the dashboard there forward. This will hopefully remove the lack of ability to be interactive. This will seem as if the dashbaord is developed for their needs particularly.}
\db{Once, the privacy issue is removed to understand more of the data, we should allow for the user to add/remove variables in to the plot as they see fit. Suggesting that the QoL variables are the ones that are added in the start, but allowing the user to remove subjects from the QoL sections will allow for the user to not only feel as if the UI is interactive, it would also create more exploration for the user.}

We also should provide much, much more user support for how to interact with and interpret the app and results. 
\db{Support for the dashboard should have a help email for towns to feel supported with interpreting the app and results. This helpful email will be connected to our research group. In which the help will not only have statisticians but also city planners that can help answer questions. As town users are looking to make changes for their towns, it could be a QoL variable that will change the status of the town, which can be addressed with the statisticians. We will be able to help them with making sure that the right variables are being used in the PCPs. }

\section{ Future Work }
Incorporating other dashboards which had slightly different focuses (99 vs 990 towns, general data display, etc.).
\db{In our research, our future work will incorporate other dashboards with general information that are slightly different focuses (99 vs 990 towns, general data display, etc.). In collaboration with DSPG at Iowa State University and the statistical modeling portions of our NSF group, we would like to transition our app to a multi-tab dashboard which allows for more specific task-oriented exploration and evaluation, rather than broad EDA and discovery.}

Discussion of the difficulty of working with researchers from other disciplines, not because of convention but because they don't know what is possible with EDA and are unfamiliar with all but goal-directed data analysis.

\db{EDA can be a difficult for a variety of groups of people, novice users and experienced researchers. While working with a variety of  disciplines because EDA can be hard to understand the use cases of exploration of data. Many experience researchers, see EDA as a way to look at data that were collected to only describe the behavior of variables that are added into models or research questions. However, EDA can be used in a data discovery, such as the work displayed in this article. Data Discovery can help with the quality of the data, in order to find new resources. EDA can help with potential casuality of town related issues. For example, if data are showing that people are no longer using public transportation, this could be due to the population shrinking but exploaring the data could suggest that more people are using ride share companies. This could as suggest that public transportation may not offer the needs for the people in the town. EDA will help identify those issues or even allow for the town leaders to look for answers.}

\section {Conclusions}
Specific takeaways - visualization research (broad) vs. dashboard design (very problem specific) is missing a critical set of research describing the specific and trying to make general suggestions as to how to proceed with the design process. 
\db{In conclusion, }

Helpful hints - showing a collection of very different apps to train collaborators on what is possible works well to empower collaborators to help brainstorm.
Test early, test often...

@@

\subsection{Visualization Revisited}
In our statistical graphs, we used Parallel Coordinate Plots (PCPs) to allow the user to look at a variety of variables, while the Guided Discovery Learning was used to makes sure that the users were informed of the information that was presented, it did not allow for the town leaders any more ease in understanding. As a result, we decided to look for easier ways to display data, such as boxplots and barplots. This direction will be limited but may be useful for the sake of ease on the town leaders.

@@

\subsection{Feedback}
We will continue to engage with town leaders to ensure that we are developing a product that will help them make better data-driven decisions that are useful to increasing their quality of life, which may lend itself to help less shrinkage in these rural areas.

### 2022-07-31 SDSS Journal Paper/Paper-Draft.tex
7.31.2022

The Smart and Connected Community (SCC) dashaboard data are primarily assembled from \url{data.iowa.gov}, with some additional datasets assembled from federal and private sources. Most of these data sets are collected at a town/city or county spatial resolution, requiring us to carefully join data to ensure that these differences are respected while collating relevant information at the city level. In addition to the more commonly available statistics derived from e.g.~the census and American Community Survey, \url{data.iowa.gov} contains several unique data sets, including local liquor sales, school building locations, town budgets and expenditures, hospital beds, Medicaid reimbursements, and other details that may provide information about local quality of life.

@@

\begin{enumerate}
\item Identifying the dimensions of rurality: population size, density, remoteness, and built-up area.
\item Selecting measureable variables to adequately represent each dimension:
    \begin{itemize}
        \item Size: logarithm of population size
        \item Density: logarithm of population density.
        \item Remoteness: network distance.
        \item Built-up area: urban area (as defined by the US Census Bureau) as a percentage of total land area.
    \end{itemize}
\item Re-scaling the variables onto bounded scales that range from 0 to 1.
\item Selecting a link function: unweighted average of the four re-scaled variable.
\end{enumerate}

@@

{\textcolor{red}{XXX Denise, can you follow up with the initial design section and explain the 2nd design and approach? Why did we change things? How did those changes impact feedback from users? E.g. wanting to show the whole range of values in the PCP vs. being visually overwhelmed. Incorporate the user feedback into the narrative. XXX}}

{\textcolor{blue}{In the initial design, we included a map for each vital service, this initially created a lag for the users' experience. As a result, we cached map directions from OSRM for each service in our database, which drastically}} {\textcolor{red}{reduced}} {\textcolor{blue}{the response time for the user. Our initial design {\textcolor{red}{did not naturally focus}} the user's eye on the most important parts of the dashboard; the redesign allowed for a cleaner {\textcolor{red}{flow}} from the top to the bottom. }}

{\textcolor{blue}{In addition to the timing due to the map loading slowly, we added the vital statistics at the county level to allow for a more robust understanding of the town and it's surroundings. The rurality index {\textcolor{red}{provided a better classification}} and the USDA sources allowed for the town to understand {\textcolor{red}{the impact of}} the closest major city {\textcolor{red}{due to commuting for work and shopping at larger stores not available within the town}}.}}

{\textcolor{red}{We also modified the parallel coordinate plots in several ways:}}

@@

{\textcolor{red}{While not all of these modifications were well received in our second round of user testing, the changes did incrementally move the dashboard display towards our goal of allowing users to explore the data and engage with it. We continued to be surprised with how well users reacted to the parallel coordinate plots, which we initially thought might be too abstract for users unfamiliar with multivariate data displays, but the ability to compare towns across multiple dimensions and examine the similarities and differences between their approaches to different services seemed to be intuitive for users once they understood that each vertical axis was a different variable.}}

@@

How did the philosophy work out? What worked and didn't (overall)? People didn't seem to like having their community facts to explore from - they already knew that information - so what should we replace it with? Collaborator suggestions included reporting capability and data stories (narratives from the communities), but neither one really inspires exploratory data analysis. What we need to do is provide enough interactivity to allow people to select interesting data sets and use that to compare. We also should provide much, much more user support for how to interact with and interpret the app and results.

@@

Incorporating other dashboards which had slightly different focuses (99 vs 990 towns, general data display, etc.). Transition to a multi-tab dashboard which allows for more specific task-oriented exploration and evaluation, rather than broad EDA and discovery.
Discussion of the difficulty of working with researchers from other disciplines, not because of convention but because they don't know what is possible with EDA and are unfamiliar with all but goal-directed data analysis.

@@

Specific takeaways - visualization research (broad) vs.~dashboard design (very problem specific) is missing a critical set of research describing the specific and trying to make general suggestions as to how to proceed with the design process.
Helpful hints - showing a collection of very different apps to train collaborators on what is possible works well to empower collaborators to help brainstorm.
Test early, test often\ldots{}

@@

In our statistical graphs, we used Parallel Coordinate Plots (PCPs) to allow the user to look at a variety of variables, while the Guided Discovery Learning was used to makes sure that the users were informed of the information that was presented, it did not allow for the town leaders any more ease in understanding. As a result, we decided to look for easier ways to display data, such as boxplots and barplots. This direction will be limited but may be useful foor the sake of ease on the town leaders.

@@

We will continue to engage with twon leaders to ensure that we are developing a product that will help them make better data-driven decisions that are useful to increasing their quality of life, which may lend itself to help less shrinkage in these rural areas.

>>>>

The Smart and Connected Community (SCC) dashboard data are primarily assembled from \url{data.iowa.gov}, with some additional datasets assembled from federal and private sources. Most of these data sets are collected at a town/city or county spatial resolution, requiring us to carefully join data to ensure that these differences are respected while collating relevant information at the city level. In addition to the more commonly available statistics derived from e.g.~the census and American Community Survey, \url{data.iowa.gov} contains several unique data sets, including local liquor sales, school building locations, town budgets and expenditures, hospital beds, Medicaid reimbursements, and other details that may provide information about local quality of life.

@@

\begin{enumerate}
\item Identifying the dimensions of rurality: population size, density, remoteness, and built-up area.
\item Selecting measurable variables to adequately represent each dimension:
    \begin{itemize}
        \item Size: logarithm of population size
        \item Density: logarithm of population density.
        \item Remoteness: network distance.
        \item Built-up area: urban area (as defined by the US Census Bureau) as a percentage of total land area.
    \end{itemize}
\item Re-scaling the variables onto bounded scales that range from 0 to 1.
\item Selecting a link function: unweighted average of the four re-scaled variable.
\end{enumerate}

@@

In the initial design, we included a map for each vital service, this initially created a lag for the users' experience. As a result, we cached map directions from OSRM for each service in our database, which drastically reduced the response time for the user. Our initial design did not naturally focus the user's eye on the most important parts of the dashboard; the redesign allowed for a cleaner flow from the top to the bottom.

In addition to the timing due to the map loading slowly, we added the vital statistics at the county level to allow for a more robust understanding of the town and it's surroundings. The rurality index provided a better classification and the USDA sources allowed for the town to understand the impact of the closest major city due to commuting for work and shopping at larger stores not available within the town.

We also modified the parallel coordinate plots in several ways:

@@

While not all of these modifications were well received in our second round of user testing, the changes did incrementally move the dashboard display towards our goal of allowing users to explore the data and engage with it. We continued to be surprised with how well users reacted to the parallel coordinate plots, which we initially thought might be too abstract for users unfamiliar with multivariate data displays, but the ability to compare towns across multiple dimensions and examine the similarities and differences between their approaches to different services seemed to be intuitive for users once they understood that each vertical axis was a different variable.

@@

How did the philosophy work out?\\
{\textcolor{blue}{The design philosophy worked well in the ways to promote guided discovery learning for the user. The blocks that were created helped create a way for the user to scroll to the next section. This framework in the design allowed for the highlighted pieces of the town in generally. This design allowed the user to see the PCP after understanding how the dashboard allowed for their town to centered. The PCP in the lower part of the dashboard allowed for the user to see the plot and adjust to the fact that they could add more towns to the plot.}}

What worked and didn't (overall)?
{\textcolor{blue}{The idea to use the guided discovery methodology created a lack of understanding in the plot for a novice user. The need to explain a visual means that the user is more likely to not use the plot effectively. Our hope was that the PCP would be a bit more "self-explanatory" than it was initially. The number of helpful hints clutters the UI and may be a distraction for a particular user. }}

People didn't seem to like having their community facts to explore from - they already knew that information - so what should we replace it with? Collaborator suggestions included reporting capability and data stories (narratives from the communities), but neither one really inspires exploratory data analysis. What we need to do is provide enough interactivity to allow people to select interesting data sets and use that to compare

{\textcolor{blue}{Many small towns don't like to see that their information is being displayed publicly or used to be compared to another city. This could a huge barrier for our resarch and work. To make sure that this does not become a problem, we could make sure that when allowing the user to compare towns in the PCP, we make sure to de-idenitify the towns in the UI. For example, the town user selects similiar population size, the each town that is populated will have a collection of ten similiar towns using a label, such as "Town 1", "Town 2", etc. }}
{\textcolor{blue}{Along with the de-idenitifying other towns within the UI, using a login system for each town will allow for the town to feel as if their data is being protected. This login system can then be used as a filter to setup the dashboard there forward. This will hopefully remove the lack of ability to be interactive. This will seem as if the dashbaord is developed for their needs particularly.}}
{\textcolor{blue}{Once, the privacy issue is removed to understand more of the data, we should allow for the user to add/remove variables in to the plot as they see fit. Suggesting that the QoL variables are the ones that are added in the start, but allowing the user to remove subjects from the QoL sections will allow for the user to not only feel as if the UI is interactive, it would also create more exploration for the user.}}

We also should provide much, much more user support for how to interact with and interpret the app and results.
{\textcolor{blue}{Support for the dashboard should have a help email for towns to feel supported with interpreting the app and results. This helpful email will be connected to our research group. In which the help will not only have statisticians but also city planners that can help answer questions. As town users are looking to make changes for their towns, it could be a QoL variable that will change the status of the town, which can be addressed with the statisticians. We will be able to help them with making sure that the right variables are being used in the PCPs. }}

@@

Incorporating other dashboards which had slightly different focuses (99 vs 990 towns, general data display, etc.).
{\textcolor{blue}{In our research, our future work will incorporate other dashboards with general information that are slightly different focuses (99 vs 990 towns, general data display, etc.). In collaboration with DSPG at Iowa State University and the statistical modeling portions of our NSF group, we would like to transition our app to a multi-tab dashboard which allows for more specific task-oriented exploration and evaluation, rather than broad EDA and discovery.}}

Discussion of the difficulty of working with researchers from other disciplines, not because of convention but because they don't know what is possible with EDA and are unfamiliar with all but goal-directed data analysis.

{\textcolor{blue}{EDA can be a difficult for a variety of groups of people, novice users and experienced researchers. While working with a variety of  disciplines because EDA can be hard to understand the use cases of exploration of data. Many experience researchers, see EDA as a way to look at data that were collected to only describe the behavior of variables that are added into models or research questions. However, EDA can be used in a data discovery, such as the work displayed in this article. Data Discovery can help with the quality of the data, in order to find new resources. EDA can help with potential casuality of town related issues. For example, if data are showing that people are no longer using public transportation, this could be due to the population shrinking but exploaring the data could suggest that more people are using ride share companies. This could as suggest that public transportation may not offer the needs for the people in the town. EDA will help identify those issues or even allow for the town leaders to look for answers.}}

@@

Specific takeaways - visualization research (broad) vs.~dashboard design (very problem specific) is missing a critical set of research describing the specific and trying to make general suggestions as to how to proceed with the design process.
{\textcolor{blue}{In conclusion, }}

Helpful hints - showing a collection of very different apps to train collaborators on what is possible works well to empower collaborators to help brainstorm.
Test early, test often\ldots{}

@@

In our statistical graphs, we used Parallel Coordinate Plots (PCPs) to allow the user to look at a variety of variables, while the Guided Discovery Learning was used to makes sure that the users were informed of the information that was presented, it did not allow for the town leaders any more ease in understanding. As a result, we decided to look for easier ways to display data, such as boxplots and barplots. This direction will be limited but may be useful for the sake of ease on the town leaders.

@@

We will continue to engage with town leaders to ensure that we are developing a product that will help them make better data-driven decisions that are useful to increasing their quality of life, which may lend itself to help less shrinkage in these rural areas.
